---
title: "Introduction to R"
author: "Roel Bakker"
date: "6-Apr-2022"
output:
  slidy_presentation: default
  beamer_presentation: default
  ioslides_presentation:
    smaller: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
*****
### Installing R

* download R from r-project.org [http://cran.at.r-project.org/]
* download RStudio Desktop from rstudio.org [http://www.rstudio.com/ide/download/]
* install both R and RStudio
* test successful installation with:
```r
demo(graphics)
```
*****
### Your first R session ....
```{r echo=TRUE}
x = rnorm(12, mean = -5)
x 
mean(x)
summary(x)
```
*****
### Draw a histogram
```{r echo=TRUE}
hist(x)
```

*****
### With more values 
```{r echo=TRUE}
hist(rnorm(1000000,mean=4),breaks=100,col="red")
```

*****
### Working on the R command-line  
  
  
* ^ v arrow keys to navigate your command history
* <- -> arrow keys to edit the command-line
* you can save the command history:

```
history()
```
* TAB for object/file completion
```
df = read.csv("data   ")
```
*****
### 1.1 Expressions 
```{r echo=TRUE}
2+3
4*5
3^(1/2)
2^8
```
R **evaluates** these expressions, i.e. determines the **value** of these expressions

*****
### 1.2 Expressions evaluating to logical values
```{r echo=TRUE}
2 < 3   
1 > 10  
5 == 5   
5 != 5
```
*****
Short for __TRUE__ and __FALSE__
```{r echo=TRUE}
T
F
```
*****
### 1.3 Variables
Every variable has a __name__ and a __type__.  

R has the following data types: __character__, __numeric__, __integer__ and __logical__.  

The name of a variable may contain 

* letters (upper/lower case matters)
* digits
* periods 
* underscores and 
* should start with a letter  

*****
Variables do not need to be declared first. 

A variable is assigned a value in an 'assignment' statement (in addition to __=__ it is possible to use __<-__ as assignment operator)

```{r echo=TRUE}
x = 200
half.x  =  x/2
half.x
y <- 12.3
y 
```
*****
### 1.3.1 character

The class() function is used to lookup the type of a variable.

```{r echo=TRUE}
firstName = "John"
class(firstName)
firstName
```

*****
### 1.3.2 numeric

```{r echo=TRUE}
heightInCm = 178.2
class(heightInCm)
heightInCm
```
*****
### 1.3.3 integer

```{r echo=TRUE}
numberOfChildren = 2
class(numberOfChildren)
numberOfChildren = 2L # number of children should be integer
class(numberOfChildren)
numberOfChildren
```
*****
### 1.3.4 logical

```{r echo=TRUE}
RisFun = T
class(RisFun)
RisFun
```
*****
### 1.3.5 Casting (coercion)

"as" can be used to cast to a different type: 
```{r echo=TRUE}
y = as.integer(3.9)
class(y)
y # the value of y is truncated
x = as.character(33)
class(x)
x
```
*****

### 1.4 Functions

A function can be called with the name of the function followed by the arguments of the function:

```{r echo=TRUE}
max(9,3,4,2,11,2)
sum(2,3,4)
prod(5,6,7)
```
*****

### 1.5 Help

```
? sum
?? product
```
*****

### 1.6 Files

* dir() shows the files in the working directory
* ls()  shows the variables defined

```
dir()
ls()
```
*****

### 2.1 Vectors: what is a vector?

A vector is an ordered collection of values of the same type

```{r echo=TRUE}
lengths = c (178.2, 180.2, 177.4) ## c for "combine"
lengths
firstNames = c ("John","Peter","James","Mary")
firstNames
```
*****

### 2.1 Vectors: vector of length 1

R works with vectors by default.    
A variable with one value is a vector of length 1.

```{r echo=TRUE}
x = c(10,20,30)
x
length(x) 
y = 33
length(y)
y[1]
```
*****
### 2.2 Sequence vectors

```{r echo=TRUE}
z = 1:10
z
z = seq(1,5)
z = seq(from=1,to=5,by=1/3)
z
9:5

```
*****
### 2.3 Vectors: selecting elements

```{r echo=TRUE}
x = 100:120
x[1]           ## the first element is element nr 1 (not 0 as in C or Python!!) 
x[c(12,14)]    ## elements 12 and 14
x[1:5]         ## elements 1 thru 5
x[-c(1,3,5,7)] ## excludes elements 1, 3, 5 and 7 
```
*****
### 2.4 Vectors: assigning values to elements
```{r echo=TRUE}
x = rep(0,10)
x [1: 3] = 2
x [5: 6] = c(5,NA)
x [7:10] = c(1,9)
x
```
*****
### 2.5 Vectors: indexing elements by name
```{r echo=TRUE}
cc = c(1200,1800,3500)
names(cc)=c("VW Polo","MG B","Rover V8")
cc["MG B"]
```
*****
### 2.6 Vectors: plotting element values

```{r echo=TRUE}
cc = c(1200,1800,3500)
names(cc)=c("VW Polo","MG B","Rover V8")
barplot(cc)
```
*****
### 2.7 Computation with vectors

```{r echo=TRUE}
a = c(1,2,3) ; a = a + 1
a
a = 1:3 ; b = 4:2 ; c = a + b
c
c/5
```
*****
### 2.8 Comparing vectors
```{r echo=TRUE}
a = c(1,2,3)
b = c(4,2,5)
a == b
```
The result is a vector of T/F values i.e. a logical vector
*****
### 2.9 Logical vectors

If a vector is compared to a single value all elements of that vector are compared; the result is a logical vector:
```{r echo=TRUE}
x = 1:5
x>2
```
*****
A logical vector can be used to select elements of the original vector:
```{r echo=TRUE}
x = 1:5
select = x>2
x[select]
```
in SQL: SELECT * FROM NUMBERS WHERE x>2;
*****
### 2.9 Logical vectors: more examples
```{r echo=TRUE}
x = 1:10
gt5 = (x > 5)
lt8 = (x < 8)
x[gt5 & lt8]
```
Use & (not &&) as 'short-circuit' boolean operators may have unexpected effects in R !!
*****
```{r echo=TRUE}
lt5 = (x < 5)
gt8 = (x > 8)
x[lt5 | gt8]
```
Use | (not ||) 

### 2.10 Applying functions to vectors
```{r echo=TRUE}
z = c(9,49,16,36)
min(z)
range(z)
sqrt(z)
sum(z)
summary(z)
```
*****
### 2.11 Often used functions with vectors
```
length()
rev()
sum(), cumsum(), prod(), cumprod()
mean(), sd(), var(), median()
min(), max(), range(), summary()
exp(), log(), sin(), cos(), tan()
round(), ceil(), 
floor(), signif()
sort(), order(), rank()
which(), which.max()
any(), all()
```
*****
### 2.12 Scatter plots
```{r echo=TRUE}
x = -5:5
y = x^2
plot(x,y)
```

*****

### 2.13 NA (Not Available)
NA indicates that a value is not available (cf NULL in SQL) 
```{r echo=TRUE}
barplot(c(25,NA,20))
```

*****

'Missing values' are denoted with NA
NA cannot be compared with other values
```{r echo=TRUE}
c(25,NA,20) == c(25,25,25)
# NA == NA # neither with itself...
is.na(NA) # the function is.na() is used to test if a value equals NA
```

*****
### 3.1 Matrices

Matrices have 2 dimensions
```{r echo=TRUE}
m1 = matrix(9,nrow=3,ncol=3)
m2 = matrix(c(11:14,21:24),nrow=2,ncol=4,byrow=T)
m2
```
*****

### 3.2 Selecting elements of a matrix

```{r echo=TRUE}
# a single element
m2[1,3]

# selecting a column 
m2[,1]

# selecting a row
m2[2,]
```
*****
### 3.3 Plotting matrix values

```{r echo=TRUE}
height=matrix(1,10,10)
height[4:6,4:6]=5
height
```
*****

```{r echo=TRUE}
height=matrix(1,10,10)
height[4:6,4:6]=5
contour(height)
```

*****

```{r echo=TRUE}
height=matrix(1,10,10)
height[4:6,4:6]=5
persp(height,expand=0.2)
```

*****

### 4.1 Calling R functions
```
? read.csv 
read.csv(file, header = TRUE, sep = ",", quote="\"", dec=".", fill = TRUE, comment.char="", ...)   
# valid ways to call a function:
read.csv("data.csv",F)            # correct
read.csv(file="data.csv",header=F)# clear
read.csv("data.csv",F,";")        # correct
read.csv(file="data.csv",header=F,sep=";") # clear
read.csv("data.csv",";")          # error!
read.csv("data.csv",sep=";")      # correct
```
*****
### 4.2 Writing R functions
```{r echo=TRUE}
mult = function(a,b){
  a*b # the same as return(a*b)
}
mult(5,6)
```
*****
### 4.3 Scope in functions
```{r echo=TRUE}
a = 1
f = function(x){
  a = 2*a
  x+a
}
f(3)
a
```
*****
```{r echo=TRUE}
g = function(x){
  x = x+3
  x
}
a=1:3
g(a)
a
```
*****
### 4.3 Scope in functions

* Functions in R don't have unexpected 'side-effects'.

* When you write a function: pretend to be living in the body of the function, 
where only data is available that is passed through the function arguments
and where you can only pass a value to the outside world through the return value...

* This means that any result of a function should be passed as a return value...

* This is not a problem as a combination of values can be passed as a list e.g. list(a,b,x)

*****
### 5.1 Data frames

Multiple (column) vectors possibly of different types, all of the same lengths
```{r echo=TRUE}
vector1 = c(178.2, 180.2, 177.4, 166.2)
vector2 =  c("John","Richard","Peter","Sophie")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)
myDataFrame
```
*****
### 5.2 Data frames - subsetting
```{r echo=TRUE}
myDataFrame[1,1:2]
myDataFrame$firstNames
```
*****
### 5.3 Data frames - logical subsetting
```{r echo=TRUE}
myDataFrame[myDataFrame$firstNames=="John",]
myDataFrame[myDataFrame$heights < 170,]
```
*****
### 6.1 Lists 

A vector with values of different types
```{r echo=TRUE}
p = list(temp=-5,precipitation="snow",sunny=T)
p
p[[1]]
p[["precipitation"]]
p$precipitation
```
* Use TAB completion!
* Use lists primarily for returning multiple types of values from a function
* Lists are not always easy to use

*****
### 7. Factors
Qualitative variables used in models
```{r echo=TRUE}
smoker = c("yes","no","yes","yes")
smokerFactor = as.factor(smoker)
smokerFactor
```
A factor contains data indexed by its 'levels' i.e. an enumeration of the values 
```{r echo=TRUE}
delay = c(1,2,2,7,3,2,3,9,1,1,2,3,6,6,5,4,32,2)
factor(delay)
table(factor(delay))
```
*****
```{r echo=TRUE}
delay = c(1,2,2,7,3,2,3,9,1,1,2,3,6,6,5,4,32,2)
barplot(table(factor(delay)))
```

*****
### 8. Summary Statistics

### 8.1 Mean

### 8.2 Median

### 8.3 Standard Deviation

### 8.4 Variance

### 8.5 Summary

*****

### 8.1 Mean

```{r echo=T}
x = c(178.2, 180.2, 177.4)
mean(x)
```
Mean is defined as:
$\bar{x} = \frac{1}{n} \sum_{i=1}^{n} x_{i}$
   
*****
### 8.2 Median

```{r echo=T}
a = c(1,2,3,4,5)
mean(a); median(a)
a[5]=20
mean(a); median(a)
```
*****
### 8.3 Standard Deviation

Standard deviation is defined as

$$ \sigma = \sqrt{\frac{ \sum (x_{i} - \bar{x})^2}{n}} $$

```{r echo=T}
a = c(1,2,3,4,5)
sd(a)
```
*****

### 8.4 Variance

Variance is defined as 

$$ \sigma^2 = \frac{ \sum (x_{i} - \bar{x})^2}{n} $$

```{r echo=T}
a = c(1,2,3,4,5)
var(a)
sd(a)==sqrt(var(a))
```
*****

### 9. Naming conventions

Variable names should be short and speficic.
Examples of often used styles:

__Camel caps__
```
myHeightCM = 188
```
__Underscore__
```
my_height_cm = 188
```
__Dot separated__
```
my.height.cm = 188
```

*****
### 10. Style guides


* [http://4dpiecharts.com/r-code-style-guide/](http://4dpiecharts.com/r-code-style-guide/)
* [http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html)
* [http://wiki.fhcrc.org/bioc/Coding_Standards](http://wiki.fhcrc.org/bioc/Coding_Standards)


*****

### 11. References

[A free R course at Cognitive Class AI (IBM)](https://cognitiveclass.ai/courses/r-101) 

[More advanced: my favorite book on Statistical Learning with R (free pdf ; Labs in R with Introduction to R in the 1st chapter)](https://www.statlearning.com/)

[Videos and slides of an excellent course accompanying Introduction to Statistical Learning (1st Ed) by Hastie and Tibshirani](https://www.dataschool.io/15-hours-of-expert-machine-learning-videos/)

[Free course at edX of of the 2nd Ed of Introduction to Statistical Learning (2nd Ed) by Hastie and Tibshirani](https://www.edx.org/course/statistical-learning)

[Slides of the course Introduction to Data Science with R by Garrett Grolemund (RStudio)](https://github.com/rstudio/Intro/tree/master/slides)

[Coursera Data Science Specialization](https://www.coursera.org/specialization/jhudatascience/1?utm_medium=listingPage)

[Data Science Specialization Community Site](http://datasciencespecialization.github.io/)

[O'Reilly Data](http://www.oreilly.com/data/)

[Try R at Code School](http://tryr.codeschool.com/)

